paper figures

### 2024-03-11 scRNAseq/paper_figures.Rmd
update figures for revision

```{r Fig_S3D,fig.width=7,fig.height=3}
RNA %>% 
  spread(gene,expression) %>% 
  dplyr::group_by(orig.ident,Patient.ID, celltype,Timepoint,Diagnosis) %>% 
  dplyr::summarise(pos=mean((ITGA4 > 0) & (ITGB7 > 0))) %>%
  dplyr::filter(Timepoint %in% c("Vedo_0d","ctrl")) %>%
  dplyr::mutate(condition=factor(ifelse(Diagnosis=='control','HC','UC'),levels=c('HC','UC'))) %>%
  ggplot(aes(x=condition,y=pos,fill=condition,color=condition)) + 
  geom_boxplot(outlier.shape=NA) +
  geom_point(size=.5,color='gray30') +
  stat_compare_means(aes(group=celltype),comparisons=list(c('UC','HC')),label = "p.format", 
                     method='wilcox.test',bracket.size=.25,tip.length=.01,size=2) +
  theme_classic() + 
  facet_grid(.~celltype,switch='both') +
  scale_fill_manual(values=condition_colors,limits=force) +
  scale_color_manual(values=condition_colors,limits=force) +
  theme(axis.text.x=element_blank(),
        legend.position='right',
        legend.key.size=unit(3.5,'mm'),
        strip.background=element_blank(),
        strip.placement='outside',
        strip.text=element_text(size=8,angle=45,hjust=1,vjust=1))  +
  labs(x='',y='fraction ITGA4/ITGB7\ndouble-positive cells',fill='',color='')
```

@@

<!-- ```{r Fig_S11F,fig.width=6,fig.height=3} -->
<!-- pl <- ggplot(df %>%  -->
<!--                dplyr::filter(Timepoint!='ctrl',celltype!='Tregs') %>% -->
<!--                gather(metric,value,-c(sample,Patient.ID,Responder,Diagnosis,Timepoint,Batch,celltype)) %>% -->
<!--                dplyr::filter(metric=='Inv.Simpson') %>% -->
<!--                dplyr::arrange(Responder,celltype,Timepoint,Patient.ID), -->
<!--              aes(x=Responder,y=value)) + -->
<!--   geom_boxplot(aes(fill=Responder,color=Responder),outlier.shape=NA) + -->
<!--   geom_point(size=1,color='gray30') + -->
<!--   theme_classic() + -->
<!--   coord_cartesian(clip='off') + -->
<!--   scale_y_log10() + -->
<!--   facet_grid(Timepoint ~ celltype,scales='free_y',switch='y') + -->
<!--   stat_compare_means(comparisons=list(c('NR','R')),label = "p.format",size=2,bracket.size=.25,tip.length=.01) + -->
<!--   scale_fill_manual(values=responder_colors,limits=force) + -->
<!--   scale_color_manual(values=responder_colors,limits=force) + -->
<!--   labs(x='',y='Inv. Simpson',title='') + -->
<!--   theme(legend.key.size=unit(4,'mm'), -->
<!--         legend.box='horizontal', -->
<!--         axis.text.x=element_blank(), -->
<!--         strip.background=element_blank(), -->
<!--         strip.text=element_text(size=8), -->
<!--         strip.placement='outside') -->

<!-- pg <- ggplotGrob(pl) -->
<!-- for(i in which(grepl("strip-t", pg$layout$name))){ -->
<!--   pg$grobs[[i]]$layout$clip <- "off" -->
<!-- } -->

<!-- grid::grid.draw(pg) -->
<!-- #pl -->
<!-- ``` -->

@@

- ITGA4 or ITGB7 positive
- Integrin-b7 positive (after determining a threshold)
- GPR15,  CCR9 or ITGAE positive

@@

cols <- c('sample','orig.ident','predicted.celltype.l2','ITGA4','ITGB7','Integrin-b7','GPR15','CCR9','ITGAE','MKI67')
pbmc_md <- do.call(rbind,lapply(pbmc[grepl('PBMC',names(pbmc))], function(x) FetchData(x,cols))) %>%
  tibble::rownames_to_column('cell') %>%
  dplyr::mutate(celltype=plyr::revalue(predicted.celltype.l2,ct_map),
                cell=paste0(sample,'_',gsub('\\.','_',cell))) %>%
  dplyr::left_join(study %>% 
                     dplyr::mutate(Patient.ID=gsub('[P]*([0-9]*)[\\/_][12]','P\\1',Characteristics.Patient.ID.),
                                   Responder=Characteristics.Responder.,
                                   Diagnosis=Characteristics.Diagnosis.,
                                   Batch=gsub(' ','',gsub('_scRNAseq','',Parameter.Value.Method.)),
                                   Timepoint=ifelse(grepl('_1_',Sample.Name),'Vedo_0d',
                                                    ifelse(grepl('_2_',Sample.Name),'Vedo_6wk','ctrl')),
                                   Material=ifelse(Characteristics.Material.=='PBMC','PBMC','CD4')) %>%
                     dplyr::select(Patient.ID,Responder,Diagnosis,Sample.Name,Material,Timepoint,Batch),
                   by=c('sample'='Sample.Name'))

@@

dfs <- list()
for (ct in c('Mmry CD4','Mmry CD8')) {
  bcs <- list(ITGA4=pbmc_md$cell[grepl(ct,pbmc_md$celltype) & (pbmc_md$ITGA4 > 0)],
              ITGB7=pbmc_md$cell[grepl(ct,pbmc_md$celltype) & (pbmc_md$ITGB7 > 0)],
              Integrin_b7=pbmc_md$cell[grepl(ct,pbmc_md$celltype) & (pbmc_md$Integrinb7=='pos')],
              GPR15=pbmc_md$cell[grepl(ct,pbmc_md$celltype) & (pbmc_md$GPR15 > 0)],
              CCR9=pbmc_md$cell[grepl(ct,pbmc_md$celltype) & (pbmc_md$CCR9 > 0)],
              ITGAE=pbmc_md$cell[grepl(ct,pbmc_md$celltype) & (pbmc_md$ITGAE > 0)],
              MKI67=pbmc_md$cell[grepl(ct,pbmc_md$celltype) & (pbmc_md$MKI67 > 0)])
  for (nn in names(bcs)) {
    if (length(bcs[[nn]]) > 0) {
      tcrcomb <- lapply(tcr_combined,
                        function(x) x[(x$barcode %in% bcs[[nn]]) & !is.na(x$TCR1) & !is.na(x$TCR2),])
      dfs[[paste0(ct,'_',nn)]] <- clonalDiversity(tcrcomb,cloneCall='gene',group='sample',exportTable=TRUE,skip.boots=TRUE) %>%
        dplyr::mutate(celltype=ct,
                      subset=nn)
    }
  }
}
df <- do.call(rbind,dfs) %>%
  dplyr::left_join(study %>% 
                     dplyr::mutate(Patient.ID=gsub('[P]*([0-9]*)[\\/_][12]','P\\1',Characteristics.Patient.ID.),
                                   Responder=Characteristics.Responder.,
                                   Diagnosis=Characteristics.Diagnosis.,
                                   Batch=gsub(' ','',gsub('_scRNAseq','',Parameter.Value.Method.)),
                                   Timepoint=ifelse(grepl('_1_',Sample.Name),'Vedo_0d',
                                                    ifelse(grepl('_2_',Sample.Name),'Vedo_6wk','ctrl'))) %>%
                     dplyr::select(Patient.ID,Responder,Diagnosis,Sample.Name,Timepoint,Batch),
                   by=c('sample'='Sample.Name')) %>%
  dplyr::arrange(Patient.ID)

for (ss in c('GPR15','Integrin_b7','ITGA4','ITGAE','ITGB7')) {
  for (ct in c('Mmry CD4','Mmry CD8')) {
    pl <- ggplot(df %>% 
                   dplyr::filter(Diagnosis=='UC') %>%
                   gather(metric,value,-c(sample,Patient.ID,Responder,Diagnosis,Timepoint,Batch,celltype,subset)) %>%
                   dplyr::filter(metric=='Inv.Simpson') %>%
                   dplyr::group_by(celltype,subset,Patient.ID) %>%
                   dplyr::mutate(ns=n()) %>%
                   dplyr::filter(ns==2,subset==ss,celltype==ct), 
                 aes(x=Timepoint,y=value)) +
      geom_boxplot(aes(fill=Timepoint,color=Timepoint),outlier.shape=NA) +
      geom_line(aes(group=Patient.ID),color='gray',size=.25) +
      geom_point(size=1,color='gray30') +
      theme_classic() +
      coord_cartesian(clip='off') +
      scale_y_log10() +
      stat_compare_means(comparisons=list(c('Vedo_6wk','Vedo_0d')),paired=TRUE,label = "p.format",size=2,bracket.size=.25,tip.length=.01) +
      scale_fill_manual(values=timepoint_colors) +
      scale_color_manual(values=timepoint_colors) +
      labs(x='',y='Inv. Simpson',title=paste0(ct,'\n',ss)) +
      theme(legend.key.size=unit(4,'mm'),
            legend.box='horizontal',
            axis.text.x=element_text(angle=90,hjust=1,vjust=0),
            strip.background=element_blank(),
            strip.placement='outside')
    print(pl)
  }
}
```

@@

<!-- ```{r Fig_S10F,fig.width=4.5,fig.height=3} -->
<!-- ggplot(df %>%  -->
<!--          dplyr::filter(Diagnosis=='UC') %>% -->
<!--          gather(metric,value,-c(sample,Patient.ID,Responder,Diagnosis,Timepoint,Batch,celltype,Material)) %>% -->
<!--          dplyr::filter(metric=='Inv.Simpson',Responder!='control') %>% -->
<!--          dplyr::mutate(Responder=ifelse(Responder=='yes','R','NR')),  -->
<!--        aes(x=Responder,y=value)) + -->
<!--   geom_boxplot(aes(fill=Responder,color=Responder),outlier.shape=NA) + -->
<!--   geom_point(size=1,color='gray30') + -->
<!--   theme_classic() + -->
<!--   coord_cartesian(clip='off') + -->
<!--   scale_y_log10() + -->
<!--   facet_grid(Timepoint~celltype,scales='free_y',switch='y') + -->
<!--   stat_compare_means(comparisons=list(c('NR','R')),label = "p.format",size=2,bracket.size=.25,tip.length=.01) + -->
<!--   scale_fill_manual(values=responder_colors) + -->
<!--   scale_color_manual(values=responder_colors) + -->
<!--   labs(x='',y='Inv. Simpson',title='') + -->
<!--   theme(legend.key.size=unit(4,'mm'), -->
<!--         legend.box='horizontal', -->
<!--         axis.text.x=element_text(angle=90,hjust=1,vjust=0), -->
<!--         strip.background=element_blank(), -->
<!--         strip.placement='outside') -->
<!-- ``` -->

@@

<!-- ```{r get_GSE73661} -->
<!-- olink_genes <- c('CXCL8','VEGFA','CD8A','CCL7','GDNF','CDCP1','CD244','IL7','TNFRSF11B','TGFB1','PLAU','IL6','IL17C','CCL2','IL17A','CXCL11', -->
<!--                  'AXIN1','TNFSF10','CXCL9','CST5','OSM','CXCL1','CCL4','CD6','SCF','IL18','SLAMF1','TGFA','CCL13','CCL11','TNFSF14','IL10RA', -->
<!--                  'FGF5','MMP1','LIFR','FGF21','CCL19','IL15RA','IL10RB','IL18R1','CD274','CXCL5','TNFSF11','HGF','IL12B','MMP10','IL10','TNF', -->
<!--                  'CCL23','CD5','CCL3','FLT3LG','CXCL6','CXCL10','EIF4EBP1','SIRT2','CCL28','DNER','S100A12','CD40','IFNG','FGF19','CCL8','CASP8', -->
<!--                  'CCL25','CX3CL1','TNFRSF9','NTF3','TNFSF12','CCL20','SULT1A1','STAMBP','ADA','LTA','CSF1') -->

<!-- meta <- read.table(file.path(path,'GSE73661','GSE73661_Vedo.txt'),header=1,sep='\t') %>% -->
<!--   dplyr::mutate(sample=make.names(geo_accession)) %>% -->
<!--   tibble::rownames_to_column("accession")  %>% -->
<!--   dplyr::mutate(diagnosis=factor(ifelse(grepl('UC',sample),'UC','control'),levels=c('control','UC')), -->
<!--                 treatment=factor(ifelse(diagnosis=='control','control', -->
<!--                                         ifelse(grepl('plac',sample),'placebo', -->
<!--                                                ifelse(grepl('IFX',sample),'IFX', -->
<!--                                                       ifelse(grepl('vdz4w',sample),'vdz4w','vdz8w')))), -->
<!--                                  levels=c('control','placebo','IFX','vdz4w','vdz8w')), -->
<!--                 timepoint=factor(ifelse(diagnosis=='control','control', -->
<!--                                         ifelse(grepl('W0',sample),'w0', -->
<!--                                                ifelse(grepl('W4|W6',sample),'w6', -->
<!--                                                       ifelse(grepl('W12',sample),'w12', -->
<!--                                                              ifelse(grepl('W52',sample),'w52','other'))))), -->
<!--                                  levels=c('control','other','w0','w6','w12','w52')), -->
<!--                 responder=ifelse(diagnosis=='control','control',ifelse(Response %in% c(0,'Placebo'),'other',Response))) %>% -->
<!--   dplyr::filter(!grepl('earlier.stop',sample), -->
<!--                 !grepl('_47$',sample), -->
<!--                 !grepl('_73$',sample), -->
<!--                 (diagnosis=='control') | ((timepoint %in% c('w0','w6')) & (treatment=='vdz4w') & (responder!='other'))) %>% -->
<!--   droplevels() -->

<!-- row.names(meta) <- meta$sample -->
<!-- colnames(meta) <- gsub('.ch1','',colnames(meta)) -->
<!-- data <- read.table(file.path(path,'GSE73661','GSE73661.txt'),header=1,sep='\t') -->
<!-- data <- data[,colnames(data) %in% row.names(meta)] -->

<!-- GSE73661 <- SummarizedExperiment(assays=list(data),colData=meta[colnames(data),]) -->
<!-- ``` -->

<!-- show a heatmap of gene expression -->

<!-- ```{r Fig_S17C,fig.width=10,fig.height=10} -->
<!-- olink_genes_here <- setdiff(olink_genes, c('SCF','TNFSF12')) -->

<!-- ha <- HeatmapAnnotation(df=colData(GSE73661)[,c('timepoint','responder','diagnosis')], -->
<!--                         show_legend=TRUE, -->
<!--                         col=list('timepoint'=c('control'='#000000', -->
<!--                                                'w0'='#ff7700', -->
<!--                                                'w6'='#882e72'), -->
<!--                                  'responder'=c('control'='#000000', -->
<!--                                                'R'='#db2F26', -->
<!--                                                'NR'='#34a048'), -->
<!--                                  'diagnosis'=c('UC'='#1965b0', -->
<!--                                                'control'='#000000'))) -->

<!-- mat <- t(scale(t(assay(GSE73661)[olink_genes_here,]))) -->
<!-- Heatmap(mat, -->
<!--         cluster_columns=TRUE, -->
<!--         show_row_names=TRUE, -->
<!--         show_column_names=FALSE, -->
<!--         top_annotation=ha, -->
<!--         column_names_gp=gpar(fontsize=6), -->
<!--         row_names_gp=gpar(fontsize=6), -->
<!--         name='z-score')  -->
<!-- ``` -->

<!-- run a few contrasts: -->

<!-- - w6vsw0 @ UC -->
<!-- - UCvscontrol -->

<!-- ```{r GSE73661_DE,fig.width=6,fig.height=5} -->
<!-- library(limma) -->
<!-- res <- list() -->

<!-- take <- colData(GSE73661)$diagnosis=='UC' -->
<!-- design <- model.matrix(~0+timepoint,data=colData(GSE73661)[take,]) -->
<!-- fit <- eBayes(lmFit(assay(GSE73661)[,take], design)) -->
<!-- contrast.matrix <- makeContrasts(w6vsw0=timepointw6-timepointw0,levels=design) -->
<!-- contrast_fit <- eBayes(contrasts.fit(fit, contrast.matrix)) -->
<!-- res[['w6vsw0@UC']] <- topTable(contrast_fit, coef='w6vsw0', n=Inf)  %>% -->
<!--   tibble::rownames_to_column('gene') %>% -->
<!--   dplyr::filter(gene %in% olink_genes)  %>% -->
<!--   dplyr::mutate(contrast='w6vsw0@UC') -->

<!-- design <- model.matrix(~0+diagnosis,data=colData(GSE73661)) -->
<!-- fit <- eBayes(lmFit(assay(GSE73661), design)) -->
<!-- contrast.matrix <- makeContrasts(UCvscontrol=diagnosisUC-diagnosiscontrol,levels=design) -->
<!-- contrast_fit <- eBayes(contrasts.fit(fit, contrast.matrix)) -->
<!-- res[['UCvscontrol']] <- topTable(contrast_fit, coef='UCvscontrol', n=Inf)  %>% -->
<!--   tibble::rownames_to_column('gene') %>% -->
<!--   dplyr::filter(gene %in% olink_genes)  %>% -->
<!--   dplyr::mutate(contrast='UCvscontrol') -->

<!-- GSE73661_DE <- do.call(rbind,res) %>% -->
<!--     dplyr::select(gene,AveExpr,logFC,P.Value,adj.P.Val,contrast) -->
<!-- ``` -->

<!-- show volcano plots (only for Olink genes!) -->

<!-- ```{r Fig_S17A_B,fig.width=4.5,fig.height=4} -->
<!-- for (ct in c('w6vsw0@UC','UCvscontrol')) { -->
<!--   res <- GSE73661_DE %>% dplyr::filter(contrast==ct) -->
<!--   pl <-  res %>% -->
<!--     ggplot(aes(x=logFC,y=-log10(P.Value))) +  -->
<!--     geom_point(aes(color=(P.Value < .05)),size=.5) +  -->
<!--     geom_text_repel(data=res %>% dplyr::filter(P.Value < .05), -->
<!--                     aes(label=gene),segment.size=.25,segment.alpha=.25,color='red') + -->
<!--     theme_classic() + -->
<!--     geom_hline(yintercept=-log10(.05),linetype='dashed',size=.25) + -->
<!--     geom_vline(xintercept=0,linetype='dashed',size=.25) + -->
<!--     scale_color_manual(values=c('black','red'),guide=FALSE) + -->
<!--     labs(title=ct) -->
<!--   print(pl) -->
<!-- } -->
<!-- ``` -->

```{r Fig_2a,fig.width=4.5,fig.height=3.5}
DimPlot(CD4.integrated,label=TRUE) +
  scale_color_brewer(palette='Set3') +
  theme_void() +
  theme(legend.key.size=unit(3,'mm'))
```

>>>>

```{r Fig_S3D,fig.width=7,fig.height=3}
pl <- RNA %>% 
  spread(gene,expression) %>% 
  dplyr::group_by(orig.ident,Patient.ID, celltype,Timepoint,Diagnosis) %>% 
  dplyr::summarise(pos=mean((ITGA4 > 0) & (ITGB7 > 0))) %>%
  dplyr::filter(Timepoint %in% c("Vedo_0d","ctrl")) %>%
  dplyr::mutate(condition=factor(ifelse(Diagnosis=='control','HC','UC'),levels=c('HC','UC'))) %>%
  ggplot(aes(x=condition,y=pos,fill=condition,color=condition)) + 
  geom_boxplot(outlier.shape=NA) +
  geom_point(size=.5,color='gray30') +
  stat_compare_means(aes(group=celltype),comparisons=list(c('UC','HC')),label = "p.format", 
                     method='wilcox.test',bracket.size=.25,tip.length=.01,size=2) +
  theme_classic() + 
  facet_grid(.~celltype,switch='both') +
  scale_fill_manual(values=condition_colors,limits=force) +
  scale_color_manual(values=condition_colors,limits=force) +
  theme(axis.text.x=element_blank(),
        legend.position='right',
        legend.key.size=unit(3.5,'mm'),
        strip.background=element_blank(),
        strip.placement='outside',
        strip.text=element_text(size=8,angle=45,hjust=1,vjust=1))  +
  labs(x='',y='fraction ITGA4/ITGB7\ndouble-positive cells',fill='',color='')

pg <- ggplotGrob(pl)
for(i in which(grepl("strip-b", pg$layout$name))){
  pg$grobs[[i]]$layout$clip <- "off"
}

@@

```{r Fig_S11F,fig.width=6,fig.height=3}
pl <- ggplot(df %>%
               dplyr::filter(Timepoint!='ctrl',celltype!='Tregs') %>%
               gather(metric,value,-c(sample,Patient.ID,Responder,Diagnosis,Timepoint,Batch,celltype)) %>%
               dplyr::filter(metric=='Inv.Simpson') %>%
               dplyr::arrange(Responder,celltype,Timepoint,Patient.ID),
             aes(x=Responder,y=value)) +
  geom_boxplot(aes(fill=Responder,color=Responder),outlier.shape=NA) +
  geom_point(size=1,color='gray30') +
  theme_classic() +
  coord_cartesian(clip='off') +
  scale_y_log10() +
  facet_grid(Timepoint ~ celltype,scales='free_y',switch='y') +
  stat_compare_means(comparisons=list(c('NR','R')),label = "p.format",size=2,bracket.size=.25,tip.length=.01) +
  scale_fill_manual(values=responder_colors,limits=force) +
  scale_color_manual(values=responder_colors,limits=force) +
  labs(x='',y='Inv. Simpson',title='') +
  theme(legend.key.size=unit(4,'mm'),
        legend.box='horizontal',
        axis.text.x=element_blank(),
        strip.background=element_blank(),
        strip.text=element_text(size=8),
        strip.placement='outside')

@@

- ITGA4, ITGB7 or ITGB1 positive
- Integrin-b7 positive (after determining a threshold)
- GPR15,  CCR9 or ITGAE positive 
- ITGA4, ITGB7 or ITGB1 negative
- Integrin-b7 negative

@@

cols <- c('sample','orig.ident','predicted.celltype.l2','ITGA4','ITGB7','ITGB1','Integrin-b7','GPR15','CCR9','ITGAE','MKI67')
pbmc_md <- do.call(rbind,lapply(pbmc[grepl('PBMC',names(pbmc))], function(x) FetchData(x,cols))) %>%
  tibble::rownames_to_column('cell') %>%
  dplyr::mutate(celltype=plyr::revalue(predicted.celltype.l2,ct_map),
                cell=paste0(sample,'_',gsub('\\.','_',cell))) %>%
  dplyr::left_join(study %>% 
                     dplyr::mutate(Patient.ID=gsub('[P]*([0-9]*)[\\/_][12]','P\\1',Characteristics.Patient.ID.),
                                   Responder=Characteristics.Responder.,
                                   Diagnosis=Characteristics.Diagnosis.,
                                   Batch=gsub(' ','',gsub('_scRNAseq','',Parameter.Value.Method.)),
                                   Timepoint=ifelse(grepl('_1_',Sample.Name),'Vedo_0d',
                                                    ifelse(grepl('_2_',Sample.Name),'Vedo_6wk','ctrl')),
                                   Material=ifelse(Characteristics.Material.=='PBMC','PBMC','CD4')) %>%
                     dplyr::select(Patient.ID,Responder,Diagnosis,Sample.Name,Material,Timepoint,Batch),
                   by=c('sample'='Sample.Name'))

@@

dfs <- list()
for (ct in c('Mmry CD4','Mmry CD8')) {
  bcs <- list(ITGA4=pbmc_md$cell[grepl(ct,pbmc_md$celltype) & (pbmc_md$ITGA4 > 0)],
              ITGB7=pbmc_md$cell[grepl(ct,pbmc_md$celltype) & (pbmc_md$ITGB7 > 0)],
              Integrin_b7=pbmc_md$cell[grepl(ct,pbmc_md$celltype) & (pbmc_md$Integrinb7=='pos')],
              GPR15=pbmc_md$cell[grepl(ct,pbmc_md$celltype) & (pbmc_md$GPR15 > 0)],
              CCR9=pbmc_md$cell[grepl(ct,pbmc_md$celltype) & (pbmc_md$CCR9 > 0)],
              ITGAE=pbmc_md$cell[grepl(ct,pbmc_md$celltype) & (pbmc_md$ITGAE > 0)],
              MKI67=pbmc_md$cell[grepl(ct,pbmc_md$celltype) & (pbmc_md$MKI67 > 0)],
              ITGB1=pbmc_md$cell[grepl(ct,pbmc_md$celltype) & (pbmc_md$ITGB1 > 0)],
              ITGA4neg=pbmc_md$cell[grepl(ct,pbmc_md$celltype) & (pbmc_md$ITGA4==0)],
              ITGB7neg=pbmc_md$cell[grepl(ct,pbmc_md$celltype) & (pbmc_md$ITGB7==0)],
              ITGB1neg=pbmc_md$cell[grepl(ct,pbmc_md$celltype) & (pbmc_md$ITGB1==0)],
              Integrin_b7_neg=pbmc_md$cell[grepl(ct,pbmc_md$celltype) & (pbmc_md$Integrinb7=='neg')])
  for (nn in names(bcs)) {
    if (length(bcs[[nn]]) > 0) {
      tcrcomb <- lapply(tcr_combined,
                        function(x) x[(x$barcode %in% bcs[[nn]]) & !is.na(x$TCR1) & !is.na(x$TCR2),])
      dfs[[paste0(ct,'_',nn)]] <- clonalDiversity(tcrcomb,cloneCall='gene',group='sample',exportTable=TRUE,skip.boots=TRUE) %>%
        dplyr::mutate(celltype=ct,
                      subset=nn)
    }
  }
}
df <- do.call(rbind,dfs) %>%
  dplyr::left_join(study %>% 
                     dplyr::mutate(Patient.ID=gsub('[P]*([0-9]*)[\\/_][12]','P\\1',Characteristics.Patient.ID.),
                                   Responder=Characteristics.Responder.,
                                   Diagnosis=Characteristics.Diagnosis.,
                                   Batch=gsub(' ','',gsub('_scRNAseq','',Parameter.Value.Method.)),
                                   Timepoint=ifelse(grepl('_1_',Sample.Name),'Vedo_0d',
                                                    ifelse(grepl('_2_',Sample.Name),'Vedo_6wk','ctrl'))) %>%
                     dplyr::select(Patient.ID,Responder,Diagnosis,Sample.Name,Timepoint,Batch),
                   by=c('sample'='Sample.Name')) %>%
  dplyr::arrange(Patient.ID)

for (ss in c('GPR15','Integrin_b7','ITGA4','ITGAE','ITGB7','ITGB1','ITGA4neg','ITGB7neg','ITGB1neg','Integrin_b7_neg')) {
  for (ct in c('Mmry CD4','Mmry CD8')) {
    pl <- ggplot(df %>% 
                   dplyr::filter(Diagnosis=='UC') %>%
                   gather(metric,value,-c(sample,Patient.ID,Responder,Diagnosis,Timepoint,Batch,celltype,subset)) %>%
                   dplyr::filter(metric=='Inv.Simpson') %>%
                   dplyr::group_by(celltype,subset,Patient.ID) %>%
                   dplyr::mutate(ns=n()) %>%
                   dplyr::filter(ns==2,subset==ss,celltype==ct), 
                 aes(x=Timepoint,y=value)) +
      geom_boxplot(aes(fill=Timepoint,color=Timepoint),outlier.shape=NA) +
      geom_line(aes(group=Patient.ID),color='gray',size=.25) +
      geom_point(size=1,color='gray30') +
      theme_classic() +
      coord_cartesian(clip='off') +
      scale_y_log10() +
      stat_compare_means(comparisons=list(c('Vedo_6wk','Vedo_0d')),paired=TRUE,label = "p.format",size=2,bracket.size=.25,tip.length=.01) +
      scale_fill_manual(values=timepoint_colors) +
      scale_color_manual(values=timepoint_colors) +
      labs(x='',y='Inv. Simpson',title=paste0(ct,'\n',ss)) +
      theme(legend.key.size=unit(4,'mm'),
            legend.box='horizontal',
            axis.text.x=element_text(angle=90,hjust=1,vjust=0),
            strip.background=element_blank(),
            strip.placement='outside')
    print(pl)
  }
}
```

@@

```{r Fig_S10F,fig.width=4.5,fig.height=3}
ggplot(df %>%
         dplyr::filter(Diagnosis=='UC') %>%
         gather(metric,value,-c(sample,Patient.ID,Responder,Diagnosis,Timepoint,Batch,celltype,Material)) %>%
         dplyr::filter(metric=='Inv.Simpson',Responder!='control') %>%
         dplyr::mutate(Responder=ifelse(Responder=='yes','R','NR')),
       aes(x=Responder,y=value)) +
  geom_boxplot(aes(fill=Responder,color=Responder),outlier.shape=NA) +
  geom_point(size=1,color='gray30') +
  theme_classic() +
  coord_cartesian(clip='off') +
  scale_y_log10() +
  facet_grid(Timepoint~celltype,scales='free_y',switch='y') +
  stat_compare_means(comparisons=list(c('NR','R')),label = "p.format",size=2,bracket.size=.25,tip.length=.01) +
  scale_fill_manual(values=responder_colors) +
  scale_color_manual(values=responder_colors) +
  labs(x='',y='Inv. Simpson',title='') +
  theme(legend.key.size=unit(4,'mm'),
        legend.box='horizontal',
        axis.text.x=element_text(angle=90,hjust=1,vjust=0),
        strip.background=element_blank(),
        strip.placement='outside')
```

@@

```{r get_GSE73661}
library(limma)
library(SummarizedExperiment)
olink_genes <- c('CXCL8','VEGFA','CD8A','CCL7','GDNF','CDCP1','CD244','IL7','TNFRSF11B','TGFB1','PLAU','IL6','IL17C','CCL2','IL17A','CXCL11',
                 'AXIN1','TNFSF10','CXCL9','CST5','OSM','CXCL1','CCL4','CD6','SCF','IL18','SLAMF1','TGFA','CCL13','CCL11','TNFSF14','IL10RA',
                 'FGF5','MMP1','LIFR','FGF21','CCL19','IL15RA','IL10RB','IL18R1','CD274','CXCL5','TNFSF11','HGF','IL12B','MMP10','IL10','TNF',
                 'CCL23','CD5','CCL3','FLT3LG','CXCL6','CXCL10','EIF4EBP1','SIRT2','CCL28','DNER','S100A12','CD40','IFNG','FGF19','CCL8','CASP8',
                 'CCL25','CX3CL1','TNFRSF9','NTF3','TNFSF12','CCL20','SULT1A1','STAMBP','ADA','LTA','CSF1')

meta <- read.table(file.path(path,'GSE73661','GSE73661_Vedo.txt'),header=1,sep='\t') %>%
  dplyr::mutate(sample=make.names(geo_accession)) %>%
  tibble::rownames_to_column("accession")  %>%
  dplyr::mutate(diagnosis=factor(ifelse(grepl('UC',sample),'UC','control'),levels=c('control','UC')),
                treatment=factor(ifelse(diagnosis=='control','control',
                                        ifelse(grepl('plac',sample),'placebo',
                                               ifelse(grepl('IFX',sample),'IFX',
                                                      ifelse(grepl('vdz4w',sample),'vdz4w','vdz8w')))),
                                 levels=c('control','placebo','IFX','vdz4w','vdz8w')),
                timepoint=factor(ifelse(diagnosis=='control','control',
                                        ifelse(grepl('W0',sample),'w0',
                                               ifelse(grepl('W4|W6',sample),'w6',
                                                      ifelse(grepl('W12',sample),'w12',
                                                             ifelse(grepl('W52',sample),'w52','other'))))),
                                 levels=c('control','other','w0','w6','w12','w52')),
                responder=ifelse(diagnosis=='control','control',ifelse(Response %in% c(0,'Placebo'),'other',Response))) %>%
  dplyr::filter(!grepl('earlier.stop',sample),
                !grepl('_47$',sample),
                !grepl('_73$',sample),
                (diagnosis=='control') | ((timepoint %in% c('w0','w6')) & (treatment=='vdz4w') & (responder!='other'))) %>%
  droplevels()

row.names(meta) <- meta$sample
colnames(meta) <- gsub('.ch1','',colnames(meta))
data <- read.table(file.path(path,'GSE73661','GSE73661.txt'),header=1,sep='\t')
data <- data[,colnames(data) %in% row.names(meta)]

GSE73661 <- SummarizedExperiment(assays=list(data),colData=meta[colnames(data),])
```

show a heatmap of gene expression

```{r Fig_S17C,fig.width=10,fig.height=10}
olink_genes_here <- setdiff(olink_genes, c('SCF','TNFSF12'))

ha <- HeatmapAnnotation(df=colData(GSE73661)[,c('timepoint','responder','diagnosis')],
                        show_legend=TRUE,
                        col=list('timepoint'=c('control'='#000000',
                                               'w0'='#ff7700',
                                               'w6'='#882e72'),
                                 'responder'=c('control'='#000000',
                                               'R'='#db2F26',
                                               'NR'='#34a048'),
                                 'diagnosis'=c('UC'='#1965b0',
                                               'control'='#000000')))

mat <- t(scale(t(assay(GSE73661)[olink_genes_here,])))
Heatmap(mat,
        cluster_columns=TRUE,
        show_row_names=TRUE,
        show_column_names=FALSE,
        top_annotation=ha,
        column_names_gp=gpar(fontsize=6),
        row_names_gp=gpar(fontsize=6),
        name='z-score')
```

run a few contrasts:

- w6vsw0 @ UC
- UCvscontrol

```{r GSE73661_DE,fig.width=6,fig.height=5}
library(limma)
res <- list()

take <- colData(GSE73661)$diagnosis=='UC'
design <- model.matrix(~0+timepoint,data=colData(GSE73661)[take,])
fit <- eBayes(lmFit(assay(GSE73661)[,take], design))
contrast.matrix <- makeContrasts(w6vsw0=timepointw6-timepointw0,levels=design)
contrast_fit <- eBayes(contrasts.fit(fit, contrast.matrix))
res[['w6vsw0@UC']] <- topTable(contrast_fit, coef='w6vsw0', n=Inf)  %>%
  tibble::rownames_to_column('gene') %>%
  dplyr::filter(gene %in% olink_genes)  %>%
  dplyr::mutate(contrast='w6vsw0@UC')

design <- model.matrix(~0+diagnosis,data=colData(GSE73661))
fit <- eBayes(lmFit(assay(GSE73661), design))
contrast.matrix <- makeContrasts(UCvscontrol=diagnosisUC-diagnosiscontrol,levels=design)
contrast_fit <- eBayes(contrasts.fit(fit, contrast.matrix))
res[['UCvscontrol']] <- topTable(contrast_fit, coef='UCvscontrol', n=Inf)  %>%
  tibble::rownames_to_column('gene') %>%
  dplyr::filter(gene %in% olink_genes)  %>%
  dplyr::mutate(contrast='UCvscontrol')

GSE73661_DE <- do.call(rbind,res) %>%
    dplyr::select(gene,AveExpr,logFC,P.Value,adj.P.Val,contrast)
```

show volcano plots (only for Olink genes!)

```{r Fig_S17A_B,fig.width=4.5,fig.height=4}
for (ct in c('w6vsw0@UC','UCvscontrol')) {
  res <- GSE73661_DE %>% dplyr::filter(contrast==ct)
  pl <-  res %>%
    ggplot(aes(x=logFC,y=-log10(P.Value))) +
    geom_point(aes(color=(P.Value < .05)),size=.5) +
    geom_text_repel(data=res %>% dplyr::filter(P.Value < .05),
                    aes(label=gene),segment.size=.25,segment.alpha=.25,color='red') +
    theme_classic() +
    geom_hline(yintercept=-log10(.05),linetype='dashed',size=.25) +
    geom_vline(xintercept=0,linetype='dashed',size=.25) +
    scale_color_manual(values=c('black','red'),guide=FALSE) +
    labs(title=ct)
  print(pl)
}
```

```{r Fig_2A,fig.width=4,fig.height=3}
DimPlot(CD4.integrated,label=TRUE) +
  scale_color_brewer(palette='Set3') +
  theme_void() +
  theme(legend.key.size=unit(3,'mm'))
```
